The imprint of global magma oceans on exoplanet demographics

### 2023-04-07 src/tex/ms.tex
edit Introduction, fix citations

\begin{note}
Recent years have brought tremendous advances in our understanding of terrestrial-sized planets, mainly due to space-based transit surveys such as \kepler\ \citep{Borucki2010} and \tess\ \citep{Ricker2014a} that pushed detection sensitivities into the terrestrial regime.
Due to the inherent detection biases of the transit technique, the currently best-probed regime of rocky exoplanets with a bulk density approaching that of Earth is that of hot or warm, close-in planets \citep{2019AREPS..47..141J,2021JGRE..12606639B}.
These bodies experience climatic conditions that are in some aspects comparable to the environment of the inner Solar System bodies at early stages of their evolution \citep{2018SSRv..214...76I,2021ChEG...81l5735C}, which sensitively affected the distribution of atmospheric volatiles between planetary core, mantle, and atmosphere.
Studies of the geophysical state and evolution of hot exoplanets can thus contribute to our understanding of the early formation stages of Earth and other habitable worlds \citep{Lichtenberg2022,2022arXiv220310056K}.

An example from the Solar System for the potential significance of these early stages are the divergent atmospheric evolutions of Venus and Earth \citep{2019JGRE..124.2015K,2021JGRE..12606643K}.
While having accreted from the same mass reservoir~\citep{Morbidelli2012} and despite their similar bulk properties, they evolved into planets with very different environmental conditions on their surfaces~\citep{Kasting1988,Hamano2013,Kane2014,Turbet2021} (Donahue et al. 1982; Bézard, B. \& de Bergh, 2007; Way et al. 2016; Way \& Del Genio 2020; ).
At the formation of the Solar System, both planets underwent a giant impact phase~\citep{2020NatGe..13..265G} that melted their mantles, leading to a magma ocean stage~\citep{2012AREPS..40..113E,2018RSPTA.37680109S}.
Magma ocean states play a substantial role in establishing the geophysical and climatic regimes of rocky planets, because atmospheric volatiles are highly soluble in magmatic liquids \citep{2021ApJ...909L..22K,Dorn2021}.
Due to these similar early formation phases, it was commonly assumed that the divergence of Venus and Earth -- in particular Venus's water loss -- occurred late in their evolution~\citep{Elkins-Tanton2013}.

However, no late desiccation needs to be invoked if a strongly infrared-absorbing, condensable species such as water is present on the planet.
Once the incident stellar flux exceeds a threshold value, a non-linear transition from mixed condensed and volatile to a pure volatile state occurs, and the resulting massive steam atmosphere limits the outgoing radiation from the (molten) planetary surface~\citep{Ingersoll1969,Kasting1988,2013NatGe...6..661G,2015ExA....40..449L}.
This runaway greenhouse state prevents the planet's rapid cooling and can extend the magma ocean stage to hundreds of \SI{}{\mega\year}, enough to remove the entire water reservoir from a rocky planet by H$_2$O photolysis and subsequent hydrodynamic escape of hydrogen. CITE
%Because hydrodynamic escape driven by stellar XUV decreases with time, the solidification timescale can become comparable to the main-sequence lifetime of the star with a sufficient water inventory \citep{2013Natur.497..607H,2015ApJ...806..216H,Lichtenberg2022}.
Venus' atmospheric composition is consistent with the planet having gone through this phase, although the past presence of a Venusian water ocean has not been definitely established~\citep{Raymond2006,Raymond2007,Hamano2013,Way2016}.

The runaway greenhouse transition is a robust prediction from climate models \citep{Forget2014,2021ApJ...919..130B,2022A&A...658A..40C}, and its impact on planetary bulk properties has been shown to be in the detectable range of current astronomical instrumentation~\citep{Goldblatt2015}.
In particular, planets in a super-runaway climate state are expected to be thermally inflated~\citep{Turbet2019,Turbet2020,Mousis2020}, which directly increases their transit radii.
However, dissolution in the magma may decrease this effect~\citep{Dorn2021}, and variations in interior chemical exchange between core and mantle material may influence the outgassed volatiles that are visible in the atmosphere via transmission spectroscopy \citep{2021ApJ...914L...4L,2022PSJ.....3..127S}.
Astronomical observations of planets that are currently in the super-runaway phase may thus constrain properties of their planetary mantles and establish an observational connection between exoplanetary interior and atmospheres~\citep{Lichtenberg2022,Wordsworth2022}.

@@

Of particular relevance is that the radiation-induced transition between a runaway greenhouse state and a temperate climate is thought to occur at a relatively sharp threshold instellation~\citep{Hamano2013}. CITE
Consequently, the instellation at which the runaway greenhouse transition occurs is aptly considered to be the inner boundary of the habitable zone, whose prevalent definition refers to the possibility of liquid water on an Earth-like planet with a CO$_2$/H$_2$O/N$_2$ atmosphere~(\citealp[][but other definitions have been put forward, e.g., ]{Kasting1993,Kopparapu2013,Kopparapu2014} \citealp{Hall2023,Spinelli2023}).
The fundamental concept of a habitable zone dates back centuries~\citep{Newton1687,Whewell1853,Shapley1953,Huang1959}, and its modern form has proven popular in the planetary literature\footnote{At the time of writing, a search of the term ``habitable zone'' in the titles and abstracts of refereed articles in NASA's Astrophysics Data System returned $\sim\num{1700}$ results.}.
Investigating its boundaries in parameter space has helped improve our understanding of habitability. CITE
Nevertheless, it should be emphasized that the habitable zone, as it stands today, is solely a concept based on theoretical predictions, and its validity continues to be debated~\citep[e.g.,][]{Cockell2016,Moore2017}.
The question naturally emerges if a habitable zone -- in its common form as boundaries in receiving instellation -- indeed exists and how the habitable zone hypothesis can be tested.
Observational tests being considered include searches for evidence of liquid water conveyed by ocean glint~\citep{Robinson2010,Lustig-Yaeger2018} or water vapor in planetary atmospheres~\citep{Suissa2020}.
A different approach relies on comparative planetology: By aiming for a statistical detection in a planet population, degeneracies and ambiguousities in making statements about the habitability of individual planets may be overcome~\citep{Checlair2019}.
Tests suggested in the literature include determining, for a range of orbital distances, atmospheric H$_{2}$O and CO$_2$ abundances, planetary albedos ~\citep{Bean2017,Bixel2021}, or colors~\citep{Crow2011}, or testing the relationship between CO$_2$ partial pressure and incident flux~\citep{Lehmer2020}.
All these tests require surveying a large enough sample of terrestrial-sized planets with next-generation instruments, rendering them out of reach in the immediate future.

@@

We organize the paper as follows:
Section~\ref{sec:met_baseline} introduces our baseline model produce synthetic star and planet samples.
    In Sect.~\ref{sec:met_rghmodel}, we describe the model component that produces runaway greenhouse-induced radius changes.
    Section~\ref{sec:met_surveys-hypotests} explains our survey simulations and hypothesis tests.
    We present our results in Sect.~\ref{sec:results} before interpreting them in Sect.~\ref{sec:discussion}.
    Finally, we summarize our findings in Sect.~\ref{sec:conclusions}.
\end{note}

@@

Hardegree-Ullman et al. (submitted) used the Gaia Catalogue of Nearby Stars \citep[hereafter GCNS,][]{Smart2021} as the basis for deriving stellar parameters for the \bioverse\ catalog.
The GCNS identified stars out to 120~pc, and includes \textit{Gaia} DR3 parallaxes and photometry in $G$, $G_{BP}$, and $G_{RP}$ bands~\citep{GaiaCollaboration2021} and $K_S$ band photometry from 2MASS~\citep{Cutri2003}, all of which were used in stellar classification.
From this information, colors and absolute magnitudes were computed, and initial color-magnitude cuts were made to remove non-main sequence stars.
Effective temperatures were computed primarily from color-temperature relations derived using the main sequence stellar parameters table from \citet{Pecaut2013}.
Next, luminosities were computed from absolute $G$-band magnitudes and a derived bolometric correction.
Stellar radii were then computed with the effective temperatures and luminosities using the Stefan-Boltzmann law, or using absolute $K_S$-band magnitudes and an empirical radius-luminosity relation from \citet{Mann2015} for targets within the absolute magnitude range of M dwarfs.
Finally, masses were derived from the mass-luminosity relation of \citet{Torres2010} for stars with $M_{\star}\gtrsim 0.7\,M_{\odot}$, and from that of \citet{Mann2019} for targets within the absolute magnitude range of M dwarfs.
The derived stellar parameters from Hardegree-Ullman et al. (submitted) were compared to measured stellar parameters for all known exoplanet hosts from the literature and were found to be consistent within 1\%, 3\%, and 5.5\% for $T_{\mathrm{eff}}$, $R_{\star}$, and $M_{\star}$, respectively, which are all below the typical measurement uncertainties of 3.3\%, 6.8\%, and 7.9\%, respectively.
From this catalog, \bioverse\ samples stars within an isotropic distance from the Solar System as required by the planetary sample size.

@@

\subsection{Orbit parameters and planet masses}\label{sec:met-orbits_masses}
Eccentric orbits alter the probability of a planet to transit \citep[e.g.,][]{Barnes2007a}.
The distribution of eccentricities $e$ of exoplanets has been found to resemble a Beta function \citep{Kipping2013b}, which we chose to draw synthetic eccentricities from.
Following~\citet{Kipping2013b}, we used a Beta distribution with parameters $a=0.867$ and $b=3.03$, and truncated the distribution at $e = 0.8$.
%Another parameter with an even higher impact on transit probabilities is the orbital inclination toward the observer $i$.
Assuming isotropic alignments of orbits, we assigned each planet an inclination drawn from a distribution uniform in $\cos(i)$.

>>>>

Recent years have brought tremendous advances in our understanding of terrestrial-sized planets, mainly due to space-based transit surveys such as \kepler\ \citep{Borucki2010} and \tess\ \citep{Ricker2014a} that pushed detection sensitivities into the terrestrial regime.
Due to the inherent detection biases of the transit technique, the currently best studied category of rocky exoplanets is that of hot or warm, close-in planets~\citep{2019AREPS..47..141J,2021JGRE..12606639B}.
These bodies experience climatic conditions that are in some aspects comparable to the environment of the inner Solar System bodies at early stages of their evolution~\citep{2018SSRv..214...76I,2021ChEG...81l5735C}, which sensitively affected the distribution of atmospheric volatiles between planetary core, mantle, and atmosphere.
Studies of the geophysical state and evolution of hot exoplanets can thus contribute to our understanding of the early formation stages of Earth and other habitable worlds~\citep{Lichtenberg2022,Krijt2022}.

An example from the Solar System for the potential significance of these early stages are the divergent atmospheric evolutions of Venus and Earth~\citep{2019JGRE..124.2015K,2021JGRE..12606643K}.
While having accreted from the same mass reservoir~\citep{Morbidelli2012} and despite their similar bulk properties, they evolved into planets with very different environmental conditions on their surfaces~\citep{Kasting1988,Hamano2013,Kane2014,Turbet2021} (Donahue et al. 1982; Bézard, B. \& de Bergh, 2007; Way et al. 2016; Way \& Del Genio 2020; ).
At the formation of the Solar System, both planets underwent a giant impact phase~\citep{2020NatGe..13..265G} that melted their mantles, leading to a magma ocean stage~\citep{2012AREPS..40..113E,2018RSPTA.37680109S}.
Magma ocean states play a substantial role in establishing the geophysical and climatic regimes of rocky planets, because atmospheric volatiles are highly soluble in magmatic liquids~\citep{2021ApJ...909L..22K,Dorn2021}.
Due to these similar early formation phases, it was commonly assumed that the divergence of Venus and Earth -- in particular Venus's water loss -- occurred late in their evolution~\citep{Elkins-Tanton2013}.

However, no late desiccation needs to be invoked if a strongly infrared-absorbing, condensable species such as water is present on the planet.
Once the incident stellar flux exceeds a threshold value, a non-linear transition from mixed condensed and volatile to a pure volatile state occurs, and the resulting massive steam atmosphere limits the outgoing radiation from the (molten) planetary surface~\citep{Ingersoll1969,Kasting1988,2013NatGe...6..661G,2015ExA....40..449L}.
This runaway greenhouse state prevents the planet's rapid cooling and can extend the magma ocean stage to hundreds of \SI{}{\mega\year} CITE, enough to remove the entire water reservoir from a rocky planet by H$_2$O photolysis and subsequent hydrodynamic escape of hydrogen.
%Because hydrodynamic escape driven by stellar XUV decreases with time, the solidification timescale can become comparable to the main-sequence lifetime of the star with a sufficient water inventory~\citep{2013Natur.497..607H,2015ApJ...806..216H,Lichtenberg2022}.
Venus' atmospheric composition is consistent with the planet having gone through this phase, although the past presence of a Venusian water ocean has not been definitely established~\citep{Raymond2006,Raymond2007,Hamano2013,Way2016}.

The runaway greenhouse transition is a robust prediction from climate models~\citep{Forget2014,Boukrouche2021,2022A&A...658A..40C}, and its impact on planetary bulk properties has been shown to be in the detectable range of current astronomical instrumentation~\citep{Goldblatt2015}.
In particular, planets in a super-runaway climate state are expected to be thermally inflated~\citep{Turbet2019,Turbet2020,Mousis2020}, which directly increases their transit radii.
However, dissolution in the magma may decrease this effect~\citep{Dorn2021}, and variations in interior chemical exchange between core and mantle material may influence the outgassed volatiles that are visible in the atmosphere via transmission spectroscopy~\citep{2021ApJ...914L...4L,2022PSJ.....3..127S}.
Astronomical observations of planets that are currently in the super-runaway phase may thus constrain properties of their planetary mantles and establish an observational connection between exoplanetary interior and atmospheres~\citep{Lichtenberg2022,Wordsworth2022}.

@@

Of particular relevance is that the radiation-induced transition between a runaway greenhouse state and a temperate climate is thought to occur at a relatively sharp threshold instellation~\citep{Hamano2013}. CITE
Consequently, the instellation at which the runaway greenhouse transition occurs is aptly considered to be the inner boundary of the habitable zone, whose prevalent definition refers to the possibility of liquid water on an Earth-like planet with a CO$_2$/H$_2$O/N$_2$ atmosphere~(\citealp[][but other definitions have been put forward, e.g., ]{Kasting1993,Kopparapu2013,Kopparapu2014} \citealp{Hall2023,Spinelli2023}).
The fundamental concept of a habitable zone dates back centuries~\citep{Newton1687,Whewell1858,Shapley1953,Huang1959}, and its modern form has proven popular in the planetary literature\footnote{At the time of writing, a search of the term ``habitable zone'' in the titles and abstracts of refereed articles in NASA's Astrophysics Data System returned $\sim\num{1700}$ results.}.
Investigating its boundaries in parameter space has helped improve our understanding of habitability. CITE
Nevertheless, it should be emphasized that the habitable zone, as it stands today, is solely a concept based on theoretical predictions, and its validity continues to be debated~\citep[e.g.,][]{Cockell2016,Moore2017}.
The question naturally emerges if a habitable zone -- in its common form as boundaries in receiving instellation -- indeed exists and how the habitable zone hypothesis can be tested.
Observational tests being considered include searches for evidence of liquid water conveyed by ocean glint~\citep{Robinson2010,Lustig-Yaeger2018} or water vapor in planetary atmospheres~\citep{Suissa2020}.
A different approach relies on comparative planetology: By aiming for a statistical detection in a planet population, degeneracies and ambiguousities in making statements about the habitability of individual planets may be overcome~\citep{Checlair2019}.
Tests suggested in the literature include determining, for a range of orbital distances, atmospheric H$_{2}$O and CO$_2$ abundances, planetary albedos ~\citep{Bean2017,Bixel2021}, or colors~\citep{Crow2011}, or testing the relationship between CO$_2$ partial pressure and incident flux~\citep{Lehmer2020}.
All these tests require surveying a large enough sample of terrestrial-sized planets with next-generation instruments, rendering them out of reach in the immediate future.

@@

We organize the paper as follows:
Section~\ref{sec:met_baseline} introduces our baseline model produce synthetic star and planet samples.
In Sect.~\ref{sec:met_rghmodel}, we describe the model component that produces runaway greenhouse-induced radius changes.
Section~\ref{sec:met_surveys-hypotests} explains our survey simulations and hypothesis tests.
We present our results in Sect.~\ref{sec:results} before interpreting them in Sect.~\ref{sec:discussion}.
Finally, we summarize our findings in Sect.~\ref{sec:conclusions}.

@@

Hardegree-Ullman et al. (submitted) used the Gaia Catalogue of Nearby Stars~\citep[hereafter GCNS,][]{Smart2021} as the basis for deriving stellar parameters for the \bioverse\ catalog.
The GCNS identified stars out to 120~pc, and includes \textit{Gaia} DR3 parallaxes and photometry in $G$, $G_{BP}$, and $G_{RP}$ bands~\citep{GaiaCollaboration2021} and $K_S$ band photometry from 2MASS~\citep{Cutri2003}, all of which were used in stellar classification.
From this information, colors and absolute magnitudes were computed, and initial color-magnitude cuts were made to remove non-main sequence stars.
Effective temperatures were computed primarily from color-temperature relations derived using the main sequence stellar parameters table from \citet{Pecaut2013}.
Next, luminosities were computed from absolute $G$-band magnitudes and a derived bolometric correction.
Stellar radii were then computed with the effective temperatures and luminosities using the Stefan-Boltzmann law, or using absolute $K_S$-band magnitudes and an empirical radius-luminosity relation from \citet{Mann2015} for targets within the absolute magnitude range of M dwarfs.
Finally, masses were derived from the mass-luminosity relation of \citet{Torres2010} for stars with $M_{\star}\gtrsim 0.7\,M_{\odot}$, and from that of \citet{Mann2019} for targets within the absolute magnitude range of M dwarfs.
The derived stellar parameters from Hardegree-Ullman et al. (submitted) were compared to measured stellar parameters for all known exoplanet hosts from the literature and were found to be consistent within 1\%, 3\%, and 5.5\% for $T_{\mathrm{eff}}$, $R_{\star}$, and $M_{\star}$, respectively, which are all below the typical measurement uncertainties of 3.3\%, 6.8\%, and 7.9\%, respectively.
From this catalog, \bioverse\ samples stars within an isotropic distance from the Solar System as required by the planetary sample size.

@@

\subsection{Orbit parameters and planet masses}\label{sec:met-orbits_masses}
Eccentric orbits alter the probability of a planet to transit~\citep[e.g.,][]{Barnes2007a}.
The distribution of eccentricities $e$ of exoplanets has been found to resemble a Beta function~\citep{Kipping2013b}, which we chose to draw synthetic eccentricities from.
Following~\citet{Kipping2013b}, we used a Beta distribution with parameters $a=0.867$ and $b=3.03$, and truncated the distribution at $e = 0.8$.
%Another parameter with an even higher impact on transit probabilities is the orbital inclination toward the observer $i$.
Assuming isotropic alignments of orbits, we assigned each planet an inclination drawn from a distribution uniform in $\cos(i)$.
